---
title: "Boxplot"
author: "Rob van der Nagel"
date: "12-3-19"
output:
  html_notebook:
    df_print: paged
    highlight: pygments
    toc: yes
    toc_float:
      collapsed: false
  html_document:
    highlight: pygments
    toc: yes
  pdf_document:
    highlight: pygments
    toc: yes
urlcolor: blue
---

```{r, include = F}
rm(list = ls())
source('lib/utils.r')
```
```{r, child = c('lib/macros.tex')}
```

# Box plot

Also known as _ mustache box_, _box diagram_ or _box & whisker plot_.
See [M&R, section 6-4] ().

`` `{r}
solar <- load.mr.mnemonic ('M&R 6-12 solar intensity Spain')
solar <- c (solar, 387)
`` `
`` `{r}
summary (solar)
`` `
`` `{r}
stripchart (solar)
`` `

One wants to merge the information from `summary` with the most important points of a 'stripchart', and at the same time abstract about the points except the outliers.

The solution is a "box plot".
The center is a dos from the $ 1 $ to the $ 3 $ the _quartile_, with the _mediaan_ as a bold line in the middle.
So half of the data is within the dos.
The _snoren_ go up to the outer points within an extra $ 1.5 $ times the interquartile distance from the dos.
* Outlier * within $ 1.5 $ to $ 3 $ times the interquartile distance from the dos are normal, outside $ 3 $ times * extreme *.
Outliers are drawn as points.
See [M&R, figure 6-13] ().

In "box plot" there is no distinction between normal and extreme outliers. Snores go up to $ 1.57 $ times the interquartile distance instead of $ 1.5 $

`` `{r}
box plot (solar)
library ('e1071')
solar skewness <- e1071 :: skewness (solar)
`` `

There are no outliers here and the graph printed above shows that the data is skewed, with a longer start after below. The _kindness_ is therefore negative and here equal to `r solar.kindness`.

# For normal distribution

`` `{r}
quartiles <- qnorm (c (1, 3) / 4)
quartiles
mustache distance <- 2 * 1.57 * 4/3
mustache distance
`` `


For a $ \ Normal {\ mu, \ sigma} $ distribution, the first quartile is at $ \ mu \ frac {2} {3} \ sigma $ and the third quartile at $ \ mu + frac {2} {3 } \ sigma $.
So the interquartile distance is equal to $ frac {4} {3} \ sigma $.
The lines are therefore $ 2 \ times1.57 \ times \ frac {4} {3} \ sigma = `r mustache.set`` sigma \ simeq 4 \ sigma $ apart.
So $ 95 $% of the data is in trouble and $ 5 $% are outliers and are drawn as points.

For a normal distribution, a box plot would be very symmetrical, with about $ 5 $% outliers (the $ 32 \ sigma $ domain).
In this sample, only $ 5 $ outliers of the expected $ 522 \ times 0.05 = 25 $ can be seen.
The number of outliers follows a $ \ Poisson {0.05} $ distribution.
`` `{r}
set.seed (20180618)
box plot (rnorm (522, 102, 20))
`` `