---
title: "Graph 1d"
author: "Rob van der Nagel"
date: "12-3-19"
output:
  html_notebook:
    df_print: paged
    highlight: pygments
    toc: yes
    toc_float:
      collapsed: false
  html_document:
    highlight: pygments
    toc: yes
  pdf_document:
    highlight: pygments
    toc: yes
urlcolor: blue
---

```{r, include = F}
rm(list = ls())
source('lib/utils.r')
```
```{r, child = c('lib/macros.tex')}
```

# Charts 1d

Given numerical data $ x_1, \ dotsc, x_n $.
We want to plot the data to gain visual insight.

`` `{r}
data <- load.mr.mnemonic ('M&R 6-31 shear strength titanium spot welds')
`` `
`` `{r}
dates
`` `

## Dotchart

Plot data point $ i $ on $ (x_i, i) $.
`` `{r}
dotchart (data)
`` `


## Tribal leaf foliage

Compare the first important decimal.
`` `{r}
voice (data)
`` `

## Stripchart or 1d scatter plot

Only useful for small samples.

`` `{r}
stripchart (data)
`` `


## Histogram

`` `{r}
hist (data)
h <- hist (data, breaks = 18)
h
`` `


## Barplot

Plot data point $ i $ on $ (i, x_i) $.
`` `{r}
bar plot (data)
`` `

Limiting a graph to a range depending on the values ​​has advantages and disadvantages.
The advantage is a better view of details and differences.
The disadvantage is whether the always visible strong differences are essential.
This problem often exists with stock prices.
`` `{r}
bar plot (data, ylim = range (data))
`` `

## Empirical distribution function

Arrange the points and plot $ (x_i, \ frac {k_i} {n}) $, where $ k_i: = | \ Set {x_j \ mid {} x_j \ le {} x_i} $, the share of smaller data points.
`` `{r}
plot (ecdf (data), main = 'Empirical distribution function')
`` `
If the sample comes from a known distribution, then the empirical distribution function is close to the real distribution function (in proportion $ n $ increases).
Here with $ \ Uniform {[1,4]} $.
`` `{r}
xrange <- seq (0.5, 4.5, by = 0.01)
plot (ecdf (runif (20, min = 1, max = 4)), main = 'Empirical distribution function')

plot (xrange, punif (xrange, min = 1, max = 4), type = 'l', col = 'black', main = 'Empirical distribution function: stochast')
n <- 10
colors <- grDevices :: rainbow (n)
for (i in 1: n) {
  lines (xrange, ecdf (runif (100, min = 1, max = 4)) (xrange), col = colors [i])
}

plot (xrange, punif (xrange, min = 1, max = 4), type = 'l', col = 'black', main = 'Empirical distribution function: convergence')
n <- 10
colors <- grDevices :: rainbow (n)
for (i in 1: n) {
  lines (xrange, ecdf (runif (2 ^ i, min = 1, max = 4)) (xrange), col = colors [i])
}
```

## Empirical probability density function

Idea: make a small error distribution around each data point (the normal distribution is popular) and add it with the weight $ \ frac {1} {n} $ to get a probability density function.


`` `{r}
plot (density (data))
`` `

Note: Can only be applied to continuous distributions. Also pay attention if you know that you have to be within a certain value domain.
Here is an example with $ \ Uniform {[1,4]} $: only the red part of the empirical probability density function is a good estimate of the real probability density function in green.

`` `{r}
simul <- runif (1000, min = 1, max = 4)
plot (density (simul), main = 'Empirical probability density function with limits')
lines (density (simul, from = 1, to = 4), col = 'red')
lines (seq (0, 5, by = 0.01), dunif (seq (0, 5, by = 0.01), min = 1, max = 4), col = 'green')
`` `