---
title: "Box plot for multiple samples"
author: "Rob van der Nagel"
date: "12-3-19"
output:
  html_notebook:
    df_print: paged
    highlight: pygments
    toc: yes
    toc_float:
      collapsed: false
  html_document:
    highlight: pygments
    toc: yes
  pdf_document:
    highlight: pygments
    toc: yes
urlcolor: blue
---

```{r reset global environment, include = F}
rm(list = ls())
```
```{r load own R methods, include = F}
source('lib/utils.r')
```
```{r tex macro load, child = c('lib/macros.tex')}
```

# Multiple independent samples

We want to graphically compare samples from two or more groups.

Given three samples
`` `{r}
x <- runif (15, min = 0, max = 2)
y <- runif (25, min = 0.1, max = 1.9) + rnorm (25, sd = 0.1)
z <- runif (12, min = -0.2, max = 2.1)
`` `
`` `{r}
summary (x)
summary (y)
summary (z)
`` `

`` `{r}
plot (density (x), col = 'red', ylim = c (0,1), ylab = 'probability', xlab = 'value', xlim = extendrange (c (x, y, z), f = 0.1 ))
lines (density (y), col = 'blue')
lines (density (z), col = 'green')
`` `

# Box plot for multiple independent samples

`` `{r}
joint <- data.frame (value = c (x, y, z), group = factor (c (rep ('x', length (x)), rep ('y', length (y)), rep ( 'z', length (z)))))
`` `

`` `{r}
joint
`` `

The formula `value ~ group` tells` boxplot` that `value` must be split according to the values ​​of the factor` group`.
`` `{r}
attach (joint)
box plot (value ~ group)
detach (joint)
`` `

Alternative:
`` `{r}
box plot (value ~ group, data = joint)
box plot (joint $ value ~ joint $ group)
`` `

# Stripchart for several independent samples

A strip chart is useful for small samples.

`` `{r}
stripchart (joint $ value ~ joint $ group)
`` `

# Quantile-quantile plot for two independent samples

See [Rmd] (./ graph_quantile.Rmd) [html] (./ graph_kwantiel.html).