---
title: "Kwantiel-kwantiel grafieken"
author: "Rob van der Nagel"
date: "19-3-2019"
output:
  html_notebook:
    df_print: paged
    highlight: pygments
    toc: yes
    toc_float:
      collapsed: false
  html_document:
    highlight: pygments
    toc: yes
  pdf_document:
    highlight: pygments
    toc: yes
urlcolor: blue
---

```{r, include = F}
rm(list = ls())
source('lib/utils.r')
```
```{r, child = c('lib/macros.tex')}
```

# Quantils

For data $ x_1, \ dotsc, x_n $ and $ p \ in [0,1] $, the * empirical $ p $ quantile * is the number $ q_p $ so that $ p $% of the $ n $ data is less than or equal than $ q_p $.

`` `{r}
solar <- load.mr.mnemonic ('M&R 6-12 solar intensity Spain')
p <- runif (1, min = 0, max = 1)
q <- quantile (solar, p)
dotchart (solar, main = sprintf ('Sun intensity with% f-quantile =% f', p, q))
lines (c (q, q), c (0, length (solar) + 1), col = 'red')
`` `

Of a continuous division $ P $ and $ p \ in [0,1] $, the * theoretical $ p $ quantile * is the number $ q_p $ so that $ P (q_p) = p $.
Or vice versa, a $ P $ divided random $ X $ meets $ P (X \ le q_p) = p $.

`` `{r}
k <- 7
p <- runif (1, min = 0, max = 1)
q <- qchisq (p, df = k)
plot.domain <- seq (0, 20, length.out = 500)
plot (plot.domain, pchisq (plot.domain, df = k)
   , type = '1', ylim = c (0.1)
   , main = sprintf ('Chi-squared (% i) with% f-quantile =% f', k, p, q)
   , xlab = "values", ylab = "chance"
   )
polygon (c (min (plot.domain), plot.domain, max (plot.domain))
      , c (0, dchisq (plot.domain, df = k), 0)
      , col = "orange", border = NA
      )
lines (c (q, q), c (0, 1), col = 'red')
lines (c (0, 20), c (p, p), col = 'green')
`` `

# Quantile-quantile graphs

Idea: plot two sets of quantiles against each other.
If the quantiles come from the same distribution, this would give a straight line.
See also [M&R, section 6-7] ().

`` `{r}
n <- 50
a <-1
b <- 2
`` `
`` `{r}
x <- runif (n, min = a, max = b)
y <- runif (2 * n, min = a, max = b)
`` `

Classical: plot twice the empirical probability density function.
`` `{r}
plot (density (x, from = a, to = b), ylim = c (0, 0.5), main = 'Empirical probability density functions')
lines (density (y, from = a, to = b), col = 'red')
`` `

Manual quantile-quantile plot.
Calculate the quantiles of each data series.
`` `{r}
x.quant <- quantile (x, probs = (1: 100) / 100)
y.kwant <- quantile (y, probs = (1: 100) / 100)
plot (x.kwant, y.kwant, type = 'p')
`` `

Better graph. The option `asp = 1` is important to keep the scale of the axes the same.
The red line is the first median ($ y = x $).
`` `{r}
plot (x.quant, y.quant, type = 'p', xlim = c (a, b), ylim = c (a, b), asp = 1)
lines (c (a, b), c (a, b), col = 'red')
`` `


Via R. The option `asp = 1` is important to keep the scale of the axes the same.
The red line is the first median ($ y = x $).
`` `{r}
qqplot (x, y, asp = 1)
lines (c (a, b), c (a, b), col = 'red')
`` `

## Data vs data

`` `{r}
n <- 250
`` `

`` `{r}
x1 <- rexp (n, rate = 1/3)
x2 <- rexp (2 * n, rate = 1/3) + 2
x3 <- 2 * x1
x4 <- x1 ^ 2
`` `

What do the probability density functions look like and what would be the effect on the quantile-quantile plots?

`` `{r}
qq plot (x1, x2, asp = 1)
domain <- range (c (x1, x2))
lines (domain, domain, col = 'red')
`` `

Add fixed values ​​= another straight line with slope $ 1 $.

`` `{r}
qq plot (x1, x3, asp = 1)
domain <- range (c (x1, x3))
lines (domain, domain, col = 'red')
`` `

With factor scales = another straight line with factor as slope.

`` `{r}
qq plot (x2, x3, asp = 1)
domain <- range (c (x2, x3))
lines (domain, domain, col = 'red')
`` `

Scale and move = linear transformation = other straight line with scale factor as slope.

`` `{r}
qq plot (x1, x4, asp = 1)
domain <- range (c (x1, x4))
lines (domain, domain, col = 'red')
`` `

Squaring has made small values ​​in $ [0,1] $ smaller and values ​​above $ 1 $ even greater.
So the point cloud is pulled apart.
But analysis is no longer possible.

## Data vs distribution

Instead of a sample, you can also use theoretical quantiles.
`` `{r}
n <- 111
x <- rpois (n, lambda = 2)
x.kwant <- quantile (x, probs = (1:50) / 50)
t.quant <- qpois ((1:50) / 50, lambda = 2)
plot (x.kwant, t.kwant, asp = 1)
lines (c (0, 10), c (0, 10), col = 'red')
`` `

Here you can also see the effect of a discrete distribution, because $ \ Poisson {2} $ - distributed randomists only take values ​​in $ \ Nat $.

## Normal distribution

We estimate the parameters of the distribution from the sample.
`` `{r}
n <- 121
m <- 2
s <- 3
x <- rnorm (n, mean = m, sd = s)
`` `
`` `{r}
x.quant <- quantile (x, probs = (1: 100) / 100)
t.kwant <- qnorm ((1: 100) / 100, mean = mean (x), sd = sd (x))
plot (x.kwant, t.kwant, asp = 1)
dom <- range (x.quant)
lines (stupid, stupid, col = 'red')
`` `

Because this is so important, it has its own command "qqnorm".
Here is the comparison with a standard normal, so the slope is not $ 1 $, but $ `rs` $.
The `qqline` command draws the slope of the normal distribution with the estimated parameters. With `asp = 1` you can visually get an idea of ​​the parameters` m` (location) and `s` (standard deviation).
`` `{r}
qq norm (x, asp = 1)
qqline (x)
`` `

Finally, if you do not have a normal distribution, you will see a deviation in the start (s) of the quantiles.
So for quantiles to $ 0 $ and / or $ 1 $.
Example: Plot $ \ Exponential {1/2} $ - divided sample against $ \ Normal {2, \ sqrt {2}} $ - distribution.
`` `{r}
x <- rexp (1234, rate = 1/2)
list (mean = mean (x), stddev = sd (x))
qq norm (x, asp = 1)
qqline (x)
`` `

With small quantiles it can be seen that the sample does not become negative, but a normal distribution also has a negative carrier.
With large quantiles, it can be seen that the sample drops more slowly (with $ e ^ {- x} $) than a normal distribution (with $ e ^ {- x ^ 2}).