---
title: "Linear regression"
author: "Rob van der Nagel"
date: "12-3-19"
output:
  html_notebook:
    df_print: paged
    highlight: pygments
    toc: yes
    toc_float:
      collapsed: false
  html_document:
    highlight: pygments
    toc: yes
  pdf_document:
    highlight: pygments
    toc: yes
urlcolor: blue
---

```{r, include = F}
rm(list = ls())
source('lib/utils.r')
```
```{r, child = c('lib/macros.tex')}
```

# Model linear regression

See also [M&R, section 11-2 and beyond] ().
We model a population of $ (Y_i) _ {i = 1} ^ n $ through

$$ Y_i = \ beta_0 + \ beta_1 x_i + E_i
   \ qquad \ text {for} i \ in \ Set {1, \ dotsc, n}
   \ ,.
$$
The $ (E_i) _ {i = 1} ^ n $ are independent and $ \ Normal {0, \ sigma} $ - divides.
Therefore $ Y_i \ DistAs \ DistAs is {\ beta_0 + \ beta_1 x_i, \ sigma} $.

The problem is to estimate $ \ beta_0 $, $ \ beta_1 $, and $ \ sigma $.
The estimation is done using the _ least-squares method and is for $ \ hat {\ beta} _0 $ in [M&R, (11-7)] (), $ \ hat {\ beta} _1 $ in [M&R, (11- 8)] () and $ \ hat {\ sigma} $ in [M&R, (11-13)] ().

Analysis of the model:

* A $ t $ key for $ H_0: \ beta_0 = 0 $ in [M&R, section 11-4] ().
* A $ t $ key for $ H_0: \ beta_1 = 0 $ in [M&R, section 11-4] ().
* A $ F $ key for $ H_0: \ beta_1 = 0 $ in [M&R, section 11-4.2] ().
* Confidence intervals for $ \ beta_0 $ and $ \ beta_1 $.
* An analysis of the residues (= _soils_) on normal distribution via quantile-quantile plot [M&R, section 11-7] () and Shapiro-Wilk test.
* The _coefficient of determination_ [M&R, section 11-7.2 and (11-34)] () which represents the proportion of the variance predicted by the model.
* A correlation test via the regression coefficient [M&R, section 11-8] ().
* Make predictions [M&R, section 11-6] ().

# Example optimum situation

`` `{r}
o.n <- sample (200: 300, 1)
o.x <- runif (o.n, 1, 5)
o.sigma <- 2
o.e <- rnorm (o.n, 0, 2)
o.beta.0 <- -2
o.beta.1 <- 3
o.y <- o.beta.0 + o.beta.1 * o.x + o.e
`` `
`` `{r}
plot (o.y ~ o.x)
`` `
`` `{r}
cor (o.x, o.y)
`` `

## Modeling with R

`` `{r}
o.lm <- lm (o.y ~ o.x)
`` `
`` `{r}
elm
`` `
`` `{r}
o.beta.hat.0 <- o.lm $ coefficients [1]
o.beta.hat.1 <- coefficients (o.lm) [2]
o.predict <- o.beta.hat.0 + o.beta.hat.1 * o.x
all (abs (o.beta.hat.0 + o.beta.hat.1 * o.x - predict (o.lm)) <. Machine $ double.eps)
`` `

`` `{r}
plot (o.y ~ o.x)
lines (o.predict ~ o.x, col = "red")
`` `

## Manual model validation

Are the residues normally distributed?
`` `{r}
o.residue <- o.y - o.predict
`` `
`` `{r}
plot (o.residue ~ o.x)
`` `
`` `{r}
qqnorm (o.residue)
qqline (o.residue)
`` `

```{r}
summary(o.residue)
shapiro.test(o.residue)
```
Sum of squares [M&R, (11-12)] ():
`` `{r}
o.sk.e <- sum (o.residue ^ 2)
o.sk.t <- sum (o.y ^ 2) - o.n * ​​mean (o.y) ^ 2
`` `

Estimator for $ \ sigma $ [M&R, (11-13)] (). There is a factor $ \ frac {1} {n-2} $ because we have already estimated two parameters $ \ beta_0 $ and $ \ beta_1 $:
`` `{r}
sigma
o.sigma.hat <- sqrt (o.sk.e / (o.n - 2))
o.sigma.hat
sd (o.residue)
sqrt (sd (o.residue) ^ 2 * (o.n - 1) / (o.n - 2))
`` `
`R ^ 2` [M&R, (11-34)] () and
`` `{r}
o.r <- 1 - o.sk.e / o.sk.t
o.r
`` `

No tests performed.

## Validation with R

All statistics and most tests are processed here.
You want that:

1. `R ^ 2` is close to $ 1 $.
2. The $ t $ keys are significant. That means that the coefficients are non-null.
3. The $ F $ test is significant.

`` `{r}
summary (o.lm)
`` `

Four graphs:

1. Residues with local estimator. You want a horizontal bar closer to the $ x $ axis.
2. Quantiel-normal plot residues. You want a straight line, especially in the start.
3. Root of absolute value of normalized residues. You want to stay within the $ [0, 3] $ interval.
4. How much residue and influence each data point has. You want a horizontal bar without outliers on the right.
`` `{r}
plot (o.lm)
`` `

Confidence intervals and forecasting
`` `{r}
confint (o.lm)
`` `
`` `{r}
o.x2 <- runif (30, -3, 10)
o.bi.c <- predict (o.lm, data.frame (o.x = o.x2), interval = "confidence")
plot (o.bi.c [, "fit"] ~ o.x2
  , pch = 'x', col = 'red'
  , main = 'Confidence intervals and simulation'
  , xlab = 'x',
  , ylab = "y")
lines (o.x2, o.bi.c [, "fit"], col = "red")
lines (o.x2, o.bi.c [, "lwr"], col = "blue")
lines (o.x2, o.bi.c [, "upr"], col = "blue")
o.bi.p <- predict (o.lm, data.frame (o.x = o.x2), interval = "prediction")
lines (o.x2, o.bi.p [, "lwr"], col = "green")
lines (o.x2, o.bi.p [, "upr"], col = "green")
points (o.x2, predict (o.lm, data.frame (ox = o.x2)) + rnorm (length (o.x2), 0, sd = summary (o.lm) $ sigma), pch = ' O')
`` `

# Example reel data

Exercises [M&R, Exercises 11-6, 11-28 and]

`` `{r}
gas <- load.mr.mnemonic ('M&R 11E-3 gasoline mileage data (in ^ 3, mpg)')
`` `

`` `{r}
str (gas)
summary (gas)
`` `
`` `{r}
plot (gas $ range ~ gas $ engine capacity)
`` `
`` `{r}
cor (gas $ range, gas $ engine capacity)
`` `

`` `{r}
gas.lm <- lm (range ~ displacement, data = gas)
`` `

`` `{r}
summary (gas.lm)
`` `

`` `{r}
plot (gas.lm)
```
`` `{r}
gas.predict <- predict (gas.lm, gas, interval = "prediction")
plot (gas.predict [, "fit"] ~ gas $ displacement
  , pch = "x"
  , col = 'red'
  , ylim = range (as.vector (gas.predict))
  , ylab = 'range'
  , xlab = 'engine capacity'
  , main = 'Confidence intervals and simulation'
  )
lines (gas $ cylinder capacity, gas.predict [, "fit"], col = "red")
lines (gas $ cylinder capacity, gas.predict [, "lwr"], col = "green")
lines (gas $ cylinder capacity, gas.predict [, "upr"], col = "green")
points (gas $ cylinder capacity, gas $ range, pch = 'o')
`` `

`` `{r}
library ('car')
car :: outlierTest (gas.lm)
`` `

Data point 20 is an outlier and strongly influences the model.

`` `{r}
gas2 <- gas [! (1: nrow (gas)% in% c (20)),]
gas2.lm <- lm (range ~ cylinder capacity, data = gas2)
summary (gas2.lm)
`` `
`` `{r}
plot (gas2.lm)
`` `
`` `{r}
gas2.predict <- predict (gas2.lm, gas2, interval = "prediction")
plot (gas2.predict [, "fit"] ~ gas2 $ displacement
  , pch = "x"
  , col = 'red'
  , ylim = range (as.vector (gas2.predict))
  )
lines (gas2 $ cylinder capacity, gas2.predict [, "fit"], col = "red")
lines (gas2 $ cylinder capacity, gas2.predict [, "lwr"], col = "green")
lines (gas2 $ cylinder capacity, gas2.predict [, "upr"], col = "green")
points (gas2 $ cylinder capacity, gas2 $ range, pch = 'o')
`` `

The influence of data point 15 is increased in the reduced data collection.