---
title: "Multilinear Regression: model"
author: "Rob van der Nagel"
date: "12-3-19"
output:
  html_notebook:
    df_print: paged
    highlight: pygments
    toc: yes
    toc_float:
      collapsed: false
  html_document:
    highlight: pygments
    toc: yes
  pdf_document:
    highlight: pygments
    toc: yes
urlcolor: blue
---

```{r, include = F}
rm(list = ls())
source('lib/utils.r')
```
```{r, child = c('lib/macros.tex')}
```

# Data

See [M&R, Table 12E-3] ().
`` `{r}
cars.df <- load.mr.mnemonic ('M&R 12E-3 Daimler 2005 cars')
`` `
`` `{r}
str (cars.df)
`` `

# Ad hoc terms

Create new columns quickly with `I (.)`. `I` may contain any expression that produces a vector of the correct length.

`` `{r}
cars.lm.1 <- lm (range ~ horsepower + weight + I (weight ^ 2) + I (weight * horsepower), data = cars.df)
summary (cars.lm.1)
`` `
Resize weight in kilograms to tons. The $ t $ keys change, but not the ANOVA.

`` `{r}
cars.lm.2 <- lm (range ~ horsepower + I (weight / 1000) + I (weight ^ 2/10 ^ 6) + I (weight / 1000 * horsepower), data = cars.df)
summary (cars.lm.2)
anova (cars.lm.2)
`` `

# Factors

`` `{r}
summary (cars.df $ transmission)
`` `

Virtual columns are created for a `factor` (nominal measurement level). Exactly one less than the number of labels in the factor. The `(Intercept)` represents the first label, and the columns indicate the difference with respect to this first label.


`` `{r}
cars.lm.3 <- lm (range ~ transmission, data = cars.df)
cars.coeff.3 <- coefficients (cars.lm.3)
summary (cars.lm.3)
anova (cars.lm.3)
`` `
Model 3 regression comparison:
$$
 \ text {range}
  = \ begin {cases}
   `r sprintf ("%. 2f ", cars.coeff.3 [1])` + E
   & \ text {if transmission = L4,}
   \\
   `r sprintf ("%. 2f ", cars.coeff.3 [1] + cars.coeff.3 [2])` + E
   & \ text {if transmission = L5,}
   \\
   `r sprintf ("%. 2f ", cars.coeff.3 [1] + cars.coeff.3 [3])` + E
   & \ text {if transmission = M6.}
  \ end {cases}
$$
Here is the $ E \ DistAs \ Normal {0, \ sigma} $ error.
This is different from first dividing the data into three groups (according to `transmission`) and building three models.
`` `{r}
cars.lm.3.L4 <- lm (range ~ 1, data = cars.df [cars.df $ transmission == 'L4',])
cars.lm.2.L5 <- lm (range ~ 1, data = cars.df [cars.df $ transmission == 'L5',])
cars.lm.2.M6 <- lm (range ~ 1, data = cars.df [cars.df $ transmission == 'M6',])
`` `

$$
  \ text {range}
  = \ begin {cases}
   `r sprintf ("%. 2f ", coefficients (cars.lm.3.L4) [1])` + E_ {L4}
   & \ text {if transmission = L4,}
   \\
   `r sprintf ("%. 2f ", coefficients (cars.lm.2.L5) [1])` + E_ {L5}
   & \ text {if transmission = L5,}
   \\
   `r sprintf ("%. 2f ", coefficients (cars.lm.2.M6) [1])` + E_ {M6}
   & \ text {if transmission = M6.}
  \ end {cases}
$$
Because here is $ E_ {L4} \ DistAs \ Normal {0, \ sigma_ {L4}} $, $ E_ {L5} \ DistAs \ Normal {0, \ sigma_ {L5}} $ and $ E_ {M6} \ DistAs \ Normal {0, \ sigma_ {M6}} $. So more leeway in the error structure.

The ANOVA shows that these virtual columns come as a package (here loss of two degrees of freedom).
There is a direct link between `cars.lm.3` and the ANOVA test on several independent samples in` cars.aov`.
`` `{r}
anova (cars.lm.3)
`` `
`` `{r}
cars.aov <- aov (cars.df $ range ~ cars.df $ transmission)
cars.aov
summary (cars.aov)
`` `

What do the virtual columns look like now? You can build this by hand.
The problem is that here the two columns in the ANOVA are considered independent of each other.

`` `{r}
cars.lm.4 <- lm (range ~ I (transmission == 'L5') + I (transmission == 'M6'), data = cars.df)
summary (cars.lm.4)
anova (cars.lm.4)
`` `

# Algebra of models in R

Overview of used expressions in the algebra for multilinear models in `R`:

* `+` adds a new term
* `:` interaction term:
     * interval measurement level: product of variables
     * nominal measurement level: fall distinction with virtual variables
     * `a: b` or` b: a` pair interaction between `a` and` b`
     * `a: b: c` triple interaction between` a`, `b` and` c` (all sequences possible)
* `*` all interactions (up to and including order of product)
* `-` removes a term
* `()` sub-model within brackets
* `^ 2` all pair interactions (often in combination with` () `)
* `1` the constant variable
* `.` stands for all variables

## Interaction with a factor

`` `{r}
cars.lm.5 <- lm (range ~ horsepower + transmission + transmission: horsepower, data = cars.df)
cars.coeff.5 <- coefficients (cars.lm.5)
summary (cars.lm.5)
`` `

This model is

$$
  \ text {range}
  = \ begin {cases}
   `r sprintf ("%. 2f ", cars.coeff.5 [1])`
   + `r sprintf ("%. 2f ", cars.coeff.5 [2])`
     \ times \ text {horsepower}
   + E
   & \ text {if transmission = L4,}
   \\
   `r sprintf ("%. 2f ", cars.coeff.5 [1] + cars.coeff.5 [3])`
   + `r sprintf ("%. 2f ", cars.coeff.5 [2] + cars.coeff.5 [5])`
     \ times \ text {horsepower}
   + E
   & \ text {if transmission = L5,}
   \\
   `r sprintf ("%. 2f ", cars.coeff.5 [1] + cars.coeff.5 [4])`
   + `r sprintf ("%. 2f ", cars.coeff.5 [2] + cars.coeff.5 [6])`
     \ times \ text {horsepower}
   + E
   & \ text {if transmission = M6.}
  \ end {cases}
$$
Again the error always has the same variance: $ E \ DistAs \ Normal {0, \ sigma} $.

Alternative spelling:
`` `{r}
summary (lm (range ~ horsepower * transmission, data = cars.df))
`` `

Without offsets for `transmission`, but always with offsets for` horsepower: transmission`. Beware, dangerous, if you don't know exactly what the linear algebra is going to do behind.
`` `{r}
summary (lm (range ~ horsepower * transmission - 1, data = cars.df))
`` `

## Other examples

Here the first order terms and the pair interactions can be seen.
The term `horsepower: displacement` is here the product between the two variables (not a factor).
`` `{r}
cars.lm.7 <- lm (range ~ (horsepower + displacement + weight) ^ 2, data = cars.df)
summary (cars.lm.7)
`` `

Here the triple interactions `horsepower: displacement: weight` is also present, but one of the three pair interactions,` displacement: weight`, has been removed.
`` `{r}
summary (lm (range ~ horsepower * displacement * weight - displacement: weight, data = cars.df))
`` `

## Extreme models

The zero model:
`` `{r}
summary (lm (range ~ 1, data = cars.df))
`` `

A fixed full model without interactions.

* `carline` is unique per line.
* `manifacturer.code` is constantly $ 20 $.
* `overdrive.code` is constantly $ 2 $.
* `air conditioning` is constantly` TRUE`.

So these variables are erased or in the model with `NAs' because there is no difference with` (Intercept) `.
`` `{r}
summary (lm (range ~. - carline - manifacturer.code - overdrive.code, data = cars.df))
`` `