---
title: "Bonferroni"
author: "Rob van der Nagel"
date: "12-3-19"
output:
  html_notebook:
    df_print: paged
    highlight: pygments
    toc: yes
    toc_float:
      collapsed: false
  html_document:
    highlight: pygments
    toc: yes
  pdf_document:
    highlight: pygments
    toc: yes
urlcolor: blue
---

```{r, include = F}
rm(list = ls())
source('lib/utils.r')
```
```{r, child = c('lib/macros.tex')}
```

# The problem of multiple keys

Suppose

* there are several different keys on the same data,
* or tests on related data (for example, various attributes in a table).
 
People want to know if they can reject all $ H_0 $ at the same time.
The problem is that each test only says something about itself, but not the (unknown) connection with other tests.

# Bonferroni inequality

The inclusion-exclusion principle
For events $ E_1, \ dotsc, E_m $ applies
$$
  \ Proba (E_1 \ cap \ dotso \ cap {} E_m)
  =
  \ sum_ {I \ subsetneq \ Set {1, \ dotsc, m}}
  (-1) ^ {\ Cardinality {I}} \ Proba \ left (\ bigcap_ {i \ in I} E_i ^ c \ right)
  \ ,.
$$
For three events $ A $, $ B $ and $ C $ it looks like this:
`` `{r, echo = F}
knitr :: include_graphics (path = './img/inclusion_exclusion.png')
`` `
Or
$$
  \ Proba (A ^ c \ cap {} B ^ c \ cap {} C ^ c)
  =
  1
  - \ Proba (A)
  - \ Proba (B)
  - \ Proba (C)
  + \ Proba (A \ cap {} B)
  + \ Proba (B \ cap {} C)
  + \ Proba (A \ cap {} C)
  - \ Proba (A \ cap {} B \ cap {} C)
  \ ,.
$$

The * Bonferroni inequality * is a short version of it (the terms with $ \ Cardinality {I} \ le 1 $) and says that

$$
  \ Proba (E_1 \ cap \ dotso \ cap {} E_m)
  \ ge
  1 - \ sum_ {i = 1} ^ m \ Proba (E_i ^ c)
$$
This inequality * ignores * all possible interactions between the events.


Suppose there are $ m $ keys with a null hypothesis $ H_0 ^ i $ and $ p $ value $ p_i $.
If $ E_i $ is the "reject $ H_0 ^ i $" event, it means that the $ p $ value of the key with
$$ H_0: \ text {must not} H_0 ^ i \ text {reject} $$
a $ p $ value with an upper limit $ \ displaystyle 1-p \ ge 1 - \ sum_ {i = 1} ^ m p_i $, or in short
$$
  p \ le \ sum_ {i = 1} ^ m p_i
  \ ,.
$$
We call this the * Bonferroni * correction.

# Do two or more tests

## Keys or expectation value within interval

This is a more general case of an * equivalence test * in [M&R, section 9-10] ().

`` `{r}
sigma <- 1
mu.really <- 5
data <- rnorm (20, mean = mu.echt, sd = sigma)
alpha <- 0.05
`` `

We want to test with $ H_0: \ mu \ le 3 $ or $ \ mu \ ge 8 $.
That is not possible at once.
We do two tests at level $ \ frac {\ alpha} {2} $.

Test 1: $ H_0: \ mu \ le 3 $ at significance level $ \ frac {\ alpha} {2} $.

`` `{r}
library ('TeachingDemos')
key below <- TeachingDemos :: z.test (data, mu = 3, sd = sigma, alternative = "greater", conf.level = 1 - alpha / 2)
key down
`` `

Press 2: $ H_0: \ mu \ ge 8 $ at level $ \ frac {\ alpha} {2} $.

`` `{r}
key above <- TeachingDemos :: z.test (data, mu = 8, sd = sigma, alternative = "less", conf.level = 1 - alpha / 2)
key above
`` `
The combined $ p $ value is the sum of the $ p $ values.
`` `{r}
p.value.with <- key. below $ p.value + key above $ p.value
p.value
`` `
So we can reject the hypothesis that $ \ mu_0 \ is in [- \ infty, 3] \ cup [8, \ infty [$.

# Comments

## Two-sided keys

A two-sided test is a combination of two one-sided tests.
Therefore $ \ frac {\ alpha} {2} $ in all formulas instead of $ \ alpha $.

## Holm correction

Given a vector of $ p $ values.
`` `{r}
ps <- sort (c (0.002166366, 0.012480863, 0.014599035, 0.024735216, 0.032798569, 0.042019270, 0.043259286))
`` `
`` `{r}
p.frame <- data.frame (p.raw = ps, p.bf = p.adjust (ps, method = 'bonferroni'), p.holm = p.adjust (ps, method = 'holm'))
p.frame
`` `

## Where multiple keys still come from

* When splitting samples and tests on the parts, the tests are independent. But Bonferroni gives an easy first approach.

* With too little data there is the * bootstrap * method. That means that you repeatedly pull from your sample to make a larger, synthetic sample.

`` `{r}
sample.small <- runif (12.0.1)
sample.small
summary (sample.small)
`` `
`` `{r}
sample.groter <- sample (sample.small, size = 500, replace = T)
summary (sample.groter)
`` `