---
title: "Mann Whitney U test"
author: "Rob van der Nagel"
date: "12-3-19"
output:
  html_notebook:
    df_print: paged
    highlight: pygments
    toc: yes
    toc_float:
      collapsed: false
  html_document:
    highlight: pygments
    toc: yes
  pdf_document:
    highlight: pygments
    toc: yes
urlcolor: blue
---

```{r, include = F}
rm(list = ls())
source('lib/utils.r')
```
```{r, child = c('lib/macros.tex')}
```

# Motivation Mann Whitney U test (two-sided)

Do two independent continuously and symmetrically distributed samples have the same expectation value?

Non-parametric alternative to the independent $ t $ test.

# Mann Whitney U theory test

See also [M&R, section 10-3] ().

I. * Assumptions *:
Provided two independent samples $ x_1, \ dotsc, x_ {n_x} $ and $ \ Normal {\ mu_1, \ sigma} $ and $ y_1, \ dotsc, y_ {n_y} $ from a continuous distribution $ P $ and $ P + \ Delta $.

II. * Greatness of interest *:
Expectation values ​​of the samples $ \ mu_1 $ and $ \ mu_2 $ and their difference $ \ Delta = mu_1 - \ mu_2 $.

III. * Hypotheses *:
Choose $ \ Delta_0 \ in \ Real $.
$$ H_0: \ mu_1 = \ mu_2 + \ Delta_0 \ quad \ text {or} \ quad \ Delta = \ Delta_0 \,. $$
$$ H_1: \ mu_1 \ not = \ mu_2 + \ Delta_0 \ quad \ text {or} \ quad \ Delta \ not = \ Delta_0 \,. $$

IV. * Test statistic *:

Here all for $ \ Delta_0 = 0 $.
Sort $ x $ and $ y $ together. Determine its position in the sorting for each value.
For $ x_i $ that is $ \ Rank (x_i) $.
List these for $ x $ and $ y $: $ w_x $ and $ w_y $.
So
$$ r_x: = \ sum_ {i = 1} ^ {n_x} \ Rank (x_i) \,. $$
$$ r_y: = \ sum_ {j = 1} ^ {n_y} \ Rank (y_j) \,. $$
Always
$$ r_x + r_y = \ sum_ {k = 1} ^ {n_x + n_y} k = \ frac {(n_x + n_y) (n_x + n_y + 1)} {2} = \ binom {n_x + n_y + 1} {2} \,. $$

The test statistic is defined differently.
In [M&R, (10-21)] (), it is
$$
  u: = \ begin {cases}
    r_x & \ text {if} n_x <n_y \ ,,
    \\
    r_y & \ text {if} n_x> n_y \ ,.
    \\
    \ frac {r_x + r_y} {2} & \ text {if} n_x = n_y \ ,.
    \ end {cases}
$$
If $ \ Delta_0 = 0 $, then we expect $ R_x $ and $ R_y $ to have somewhat the same distribution. So we expect that $ U $ is close to $ \ frac {(n + m) (n + m + 1)} {4} $.
Small values ​​of $ U $ let us reject $ H_0 $.
So it is basically a left-sided test.
The $ p $ value is taken from [M&R, Appendix A, Table X] ().

For `R` it is
$$ h: = r_x - \ binom {n_x + 1} {2} \,. $$
Here, small and large values ​​of $ U $ let us reject $ H_0 $.
So it is basically a two-way test.
The $ p $ value is obtained from `pwilcox (., $ N_x $, $ n_y $)`.
Under $ H_0 $, $ U $ has a symmetrical distribution on $ \ Set {0, \ dotsc, \ binom {n_x + n_y + 1} {2} - \ binom {n_y + 1} {2}} $.

V. * Decision *:
Choose significance level $ \ alpha \ in [0,1] $.
If $ p \ ge {} \ alpha $, then you must continue to accept $ H_0: \ mu_1 \ ge \ mu_2 + \ Delta_0 $.
If $ p \ le {} \ alpha $, you can reject $ H_0: \ mu_1 \ ge \ mu_2 + \ Delta_0 $ and you know that $ H_1: \ mu_1 <\ mu_2 + \ Delta_0 $ at significance level $ \ alpha $ applies to.

VI. *Good to know*:
It is a kind of Wilcoxcon test for two independent samples.
Non-parametric test.

# Example Mann Whitney U

See [M&R, problem 10-31] ().
`` `{r}
circuits <- load.mr.mnemonic ('M&R 10-31 current in μA')
`` `
``` {r}
str (circuits)
summary (circuits)
`` `

`` `{r}
circuit.1 <- circuits $ circuit.1
circuit.2 <- na.omit (circuits $ circuit.2)
circuit.all <- c (circuit.1, circuit.2)
n.1 <- length (circuit.1)
n.2 <- length (circuit.2)
n.all <- n.1 + n.2
`` `


`` `{r}
xlims <- extendrange (r = range (circuit.all), f = 0.5)
plot (density (circuit.1), xlim = xlims, main = "empirical densities")
lines (density (circuit.2), col = 'red')
`` `

We test:
$$ H_0: \ mu_1 = \ mu_2 \ quad \ text {or} \ quad \ Delta = 0 \.
$$ H_1: \ mu_1> \ mu_2 \ quad \ text {or} \ quad \ Delta> 0 \.
Manual:
`` `{r}
ranks <- rank (circuit.all)
grades
r.1 <- sum (grades [1: n.1])
r.1
r.2 <- sum (grades [(n.1 + 1): n.all])
r.2
`` `
Such as [M&R, (10-21)] ():
`` `{r}
u.mr <- r.1 * (n.1> = n.2) + r.2 * (n.1 <n.2)
u.mr
`` `
From the table at $ (9.8) $ we get that `u.mr` for $ \ alpha = 0.01 $ smaller or equal to $ 45 $ and for $ \ alpha = 0.05 $ smaller or equal to $ 51 $.

Such as "R":
`` `{r}
correction.r <- choose (n.1 + 1, 2)
u.r <- r.1 - correction.r
u.r
pwilcox (u.r, n.1, n.2, lower.tail = F)
`` `
Automatically with "R".
The $ p $ value is slightly different, because "R" makes a correction because of "ties".
`` `{r}
alpha <- 0.01
wilcox.test (circuits $ circuit.1, circuits $ circuit.2, alternative = "greater", paired = F, conf.int = T, conf.level = 1 - alpha)
`` `
We cannot reject $ H_0 $ at significance level $ `r alpha` $.

For comparison, a Welch test (independent t test with unequal variances).
`` `{r}
shapiro.test (circuits $ circuit.1)
shapiro.test (circuits $ circuit.2)
t.test (circuits $ circuit.1, circuits $ circuit.2, alternative = "greater")
`` `