---
title: "Pearson test"
author: "Rob van der Nagel"
date: "12-3-19"
output:
  html_notebook:
    df_print: paged
    highlight: pygments
    toc: yes
    toc_float: true
  html_document:
    highlight: pygments
    toc: yes
  pdf_document:
    highlight: pygments
    toc: yes
urlcolor: blue
---

```{r, include = F}
rm(list = ls())
source('lib/utils.r')
```
```{r, child = c('lib/macros.tex')}
```

# Pearson key

Problem: Sample from discrete distribution. Does this sample come from a given probability distribution $ P $? This is an adjustment test.

## Theory

We follow the 7 steps in [M&R, section 9-1.6] ().
Contents of [M&R, section 9-7] ().

0. * Assumptions *:
$ P $ is a probability distribution on $ \ Set {1, \ dotsc, l} $. Data $ x_1, \ dotsc, x_n $ is a sample of i.i.d. $ P $ distributed probability variables.

1. * Parameter of interest *: Distribution $ P $.

2. * Null hypothesis *:
For a fixed probability distribution $ P_0 $:
$$ H_0: P = P_0

3. * Alternative hypothesis *:
$$ H_1: P \ not = P_0 \

4. * Test variable *:

The * empirical frequency * counts how many data are in each course:
$$
  f:
  \ qquad
  \ Set {1, \ dotsc, l} \ to \ Nat
  \ qquad
  i \ mapsto {} \ Cardinality {\ Set {1 \ le j \ le n \ mid x_j = i}}
  \ ,.
$$
The * theoretical frequency * counts how much data we expect in each course:
$$
  F_0:
  \ qquad
  \ Set {1, \ dotsc, l} \ to \ Nat
  \ qquad
  i \ mapsto {} nP_0 (i)
  \ ,.
$$
We compare the empirical with the theoretical frequency.
The test statistic is
$$
  k ^ 2: = \ sum_ {i = 1} ^ n \ frac {(f (i) -F_0 (i)) ^ 2} {F_0 (i)}
  \ ,.
$$
That is a kind of $ 1 ^ 2 $ norm (euclicidic distance) between $ f $ and $ F_0 $.
For sufficiently large $ n $, $ K ^ 2 $ has a $ \ ChiSquare {l-1} $ - distribution.
This is a consequence of the central limit.
In practice, the central limit theorem requires that $ F_0 (i) $ should not be too small (rule of thumb: $ F_0 {} \ ge 5 $ for each course).
Otherwise, courses (= values) must be summarized.
The parameter is $ n-1 $, because $ f (l) = n - \ sum_ {i = 1} ^ {l-1} f (i) $ eliminates a degree of freedom.
If parameters are needed for $ P_0 $, then a further degree of freedom is deducted for each parameter.

5. * Decision *:
Choose a significance level $ \ alpha \ in [0,1] $.
Reject $ H_0 $ as the $ p $ value $ \ le \ alpha $.

6. * Calculation *:
If $ K ^ 2 \ DistAs \ ChiSquare {k-1} $, the $ p $ value is calculated by
$$ p: = \ Proba (k ^ 2 \ le K ^ 2) \,. $$
See also [M&R, Appendix A, Table IV] ().

7. * Conclusion *:
If $ p \ ge {} \ alpha $, then you must continue to accept $ H_0 $.
If $ p \ le {} \ alpha $, you can reject $ H_0 $.

## Data

We simulate 100 throws of a die.
`` `{r}
n <- 100
1 <- 6
throws <- sample (1: l, size = n, replace = T)
`` `
`` `{r}
freq.emp <- table (throws)
freq.emp
bar plot (freq.emp)
`` `
`` `{r}
freq.theo <- rep (n / l, l)
bar plot (freq.emp)
lines (c (0, 1 + 1), c (n / l, n / l), col = 'red')
`` `

## Manual

`` `{r}
all (freq.theo> = 5)
degrees of freedom <- 1 - 1
k.quared <- sum ((freq.emp - freq.theo) ^ 2 / freq.theo)
squared
p.value <- pchisq (k.quartered, df = degrees of freedom, lower.tail = F)
p.value
`` `
`` `{r}
alpha <- 0.1
p.value <= alpha
`` `

_Conclusion_: We must continue to assume that the dice was fair.

## Calculation with R

Do not forget the `p =` and the option `rescale.p = T`.
Otherwise it becomes a different test which we will do later.
`` `{r}
chisq.test (freq.emp, p = freq.theo, rescale.p = T)
`` `
\ # With parameter estimate


`` `{r}
sample <- load.mr.mnemonic ('M&R 9-101 vehicle count / minute frequencies')
str (sample)
summary (sample)
sample
hist (sample $ count)
plot (sample)
`` `
Because the `count` all added data in boxes, its sum counts the size of the sample (we have not seen this data!).
`` `{r}
n <- sum (sample $ count)
`` `
The average must be calculated per hand, because we only have a sample of the sample.
`` `{r}
m <- sum (sample $ vehicles * sample $ count) / n
m
plot (0: 100, dpois (0: 100, lambda = m) * n)
points (sample $ vehicles, sample $ count, col = 'red')
`` `

`` `{r}
vehicles.range <- range (sample $ vehicles)
vehicles.range
boxes <- vehicles.range [1]: vehicles.range [2]
dpois (boxes, lambda = m) * n> = 5
sum (dpois (boxes, lambda = m))
`` `

The problem: only `r sum (dpois (courses, lambda = m))` of the theoretical probability lies in these courses.
`` `{r}
ppois (min (boxes) - 1, lambda = m)
ppois (max (boxes), lambda = m, lower.tail = F)
`` `

We create a new course on the left and add it to the last on the right.
```{r}
compartments.groter <- (vehicles.range [1] -1) :( vehicles.range [2])
k <- length (subjects.larger)
freq.theo.groter <- n * c (
  ppois (min (boxes) - 1, lambda = m)
  , dpois (boxes, lambda = m)
)
l <- length (boxes)
freq.theo.groter [l + 1] <- freq.theo.groter [l + 1] + ppois (max (subjects.groter), lambda = m, lower.tail = F)
freq.emp.groter <- c (0, sample $ count)
`` `

Check whether new courses are good.
`` `{r}
all (freq.theo.groter> = 5)
`` `
`` `{r}
chi.kwadraat <- sum ((freq.emp.groter - freq.theo.groter) ^ 2 / freq.theo.groter)
chi. squared
p.value <- pchisq (chi. square, df = k - 1 - 1, lower.tail = F)
p.value
`` `

_Conclusion_: We can reject that this was a $ \ Poisson {`r m`} $ distribution.

# Continuous distributions

A perfect sample disturbed by a small Gaussian noise.
`` `{r}
n = 1000
x <- rgamma (n, shape = 10, scale = 3) + rnorm (n, mean = 0, sd = .1)
`` `
`` `{r}
h <- hist (x)
h
`` `
The histogram already divides the data into sections.
Handy to use.
`` `{r}
str (h)
`` `
The extra $ 0 $ and $ 1 $ for the vectors are for the odds in the left and right boxes.
`` `{r}
k <- length (h $ counts)
k
proba.midbreaks <- pgamma (h $ breaks [2: k], shape = 10, scale = 3)
proba.theo <- c (proba.midbreaks, 1) - c (0, proba.midbreaks)
freq.theo <- n * proba.theo
`` `
The midpoints `h $ mids` we have the grace of the histogram.
`` `{r}
plot (h $ mids, freq.theo)
points (h $ mids, h $ counts, col = 'red')
`` `
The test.
`` `{r}
chisq.test (h $ counts, p = freq.theo, rescale.p = T)
`` `
There is a warning.
We check and, yes, three courses do not meet the requirements.
`` `{r}
freq.theo> = 5
`` `