---
title: "Wilcoxon test"
author: "Rob van der Nagel"
date: "12-3-19"
output:
  html_notebook:
    df_print: paged
    highlight: pygments
    toc: yes
    toc_float:
      collapsed: false
  html_document:
    highlight: pygments
    toc: yes
  pdf_document:
    highlight: pygments
    toc: yes
urlcolor: blue
---

```{r, include = F}
rm(list = ls())
source('lib/utils.r')
```
```{r, child = c('lib/macros.tex')}
```

# Wilcoxon key

If $ P $ is a symmetrical distribution, the median is equal to the expected value.
Use the median to estimate the expected value.

## Theory test (right-sided)

We follow the 7 steps in [M&R, section 9-1.6] ().
See also [M&R, section 9-9.2] ().

I. * Assumptions *:
Sample $ x_1, \ dotsc, x_n $ from symmetrical continuous distribution $ P $.

Note: Suppose the median is $ \ mu $. Symmetry can be tested by testing $ \ Set {x_1, \ dotsc, x_n} $ vs $ \ Set {\ mu-x_1, \ dotsc, \ mu-x_n} $ for the same distribution.

II. * Greatness of importance *: $ \ mu $ expected value of $ P $. This does not have to be a parameter of $ P $.

III. * Hypotheses *: For a fixed value $ \ mu_0 $.

$$ H_0: \ mu \ le \ mu_0 \,. $$
$$ H_1: \ mu> \ mu_0 \

IV. * Test statistic *:

We calculate with "$ \ mu = \ mu_0 $".
Suppose
$$ y_i: = x_i - \ mu_0 \,. $$

Sort their absolute values: $ \ Modulus {y_i} $ has position $ \ Rank (\ Modulus {y_i}) \ in \ Set {1, \ dotsc, n} $.
Their weighted sum is
$$
  w: = \ sum_ {i = 1} ^ n \ Signum (y_i) \ Rank (\ Modulus {y_i})
  \ ,.
$$
The $ p $ value is taken from [M&R, Appendix A, Table IX] ().

V. * Decision *:
Choose significance level $ \ alpha \ in [0,1] $.
If $ p \ ge {} \ alpha $, then you must continue to accept $ H_0: \ mu \ le \ mu_0 $.
If $ p \ le {} \ alpha $, you can reject $ H_0: \ mu \ le \ mu_0 $ and know that $ H_1: \ mu> \ mu_0 $ applies at level $ \ alpha $.

VI. *Good to know*:
Non-parametric test.
Operates at ordinal measurement level using the median.
Problems with non-continuous (so all discrete) distribution, because then grades do not have to be unambiguous.
Distribution of the test statistics in "R" with "psignrank".

## Issue

`` `{r}
drying time data <- load.mr.mnemonic ('M&R 9-117 paint drying time in h')
`` `
`` `{r}
drying time
summary (drying time.data)
`` `
`` `{r}
plot (density (drying time.data))
`` `

`` `{r}
alpha <- 0.08 # chosen by yourself
mu.zero <- 1.5
`` `

$ H_0: \ mu \ le `r mu.zero` $ vs $ H_1: \ mu>` r mu.zero` $.

## Calculation step by step

`` `{r}
y <- drying time.data - mu.zero
y
z <- abs (y)
z
```
`` `{r}
data.frame (abs. difference = z, rank = rank (z))
`` `

Because the precision of the data is so low, there are equal values ​​in the 'rank' column of the table above.

`` `{r}
w <- sum (rank (z) * sign (y))
w
`` `

Because in [M&R, Appendix A, Table IX] () there is a value of $ 60 $ at $ n = 20 $ and $ \ alpha = 0.05 / 0.10 $ (we have $ \ alpha = `r alpha` $), that $ w $ is much larger and therefore we can reject $ H_0 $.

## Calculation with R

`` `{r}
key <- wilcox.test (x = drying time.data, mu = mu.zero, alternative = 'greater', conf.int = TRUE, conf.level = 1 - alpha)
test
`` `

The error message comes from the "ties", where the rank in $ \ Modulus {y} $ is not unambiguous.

Yes, we are pretty sure that $ \ mu \ ge`r mu.zero` $, because the $ p $ value is $ `r key $ p.value` $.

# Not the t test?

Can't we use the $ t $ key? We first test graphically and numerically on a normal distribution.

`` `{r}
qq standard (drying time.data)
qqline (drying time.data)
`` `
`` `{r}
shapiro.test (drying time.data)
`` `

`` `{r}
t.test (x = drying time.data, mu = mu.zero, alternative = 'greater', conf.int = TRUE, conf.level = 1 - alpha)
`` `

