---
title: "The t-test"
author: "Rob van der Nagel"
date: "12-3-19"
output:
  html_notebook:
    df_print: paged
    highlight: pygments
    toc: yes
    toc_float:
      collapsed: false
  html_document:
    highlight: pygments
    toc: yes
  pdf_document:
    highlight: pygments
    toc: yes
urlcolor: blue
---

```{r, include = F}
rm(list = ls())
source('lib/utils.r')
```
```{r, child = c('lib/macros.tex')}
```

# Student t distribution

Extract from [M&R, section 8-2.1] ().
If $ (X_i) _ {i = 1} ^ n $ i.i.d. $ \ Normal {\ mu, \ sigma} $ divided, then the average
$$ \ Mean {X}: = \ frac {1} {n} \ sum_ {i = 1} ^ n X_i $$
and the standard deviation
$$ S: = \ frac {1} {n-1} \ sum_ {i = 1} ^ n (X_i- \ Mean {X}) ^ 2 \,. $$
The stochast
$$ T: = \ frac {\ Mean {X} - \ mu} {S / \ sqrt {n}} $$
$ \ TDist {n-1} $ is divided with $ (n-1) $ degrees of freedom.

`` `{r}
a <- -5
b <- 5
n <- 200
points <- (0: n) / n * (b - a) + a
plot (0, ylim = c (0,0.5), xlim = range (points), xlab = 'value', ylab = 'probability density', main = 'Probability density of different t-distributions')

degrees of freedom <- c (1,2,3,4,5,10,20,30,40,50,75,100)
colors <- grDevices :: rainbow (length (degrees of freedom))
for (i in 1: length (degrees of freedom)) {
  lines (dt (points, df = degrees of freedom [i]) ~ points, col = colors [i])
}
`` `
`` `{r, echo = F}
library (plotrix)
`` `

Color "` r color.id (colors [1]) [1] `" is $ \ TDist {`r degrees of freedom [1]`} $ up to "` r color.id (colors [length (degrees of freedom)]) [1] `" is $ \ TDist {`r degrees of freedom [length (degrees of freedom)]`} $.
With $ \ infty $ degrees of freedom, $ \ TDist {n} $ is approaching $ \ Normal {0.1} $.
The most spread is $ \ TDist {1} $ (the _Cauchy distribution_), with larger starts.

# Double-sided t-test

## Theory (two-way test)

We follow the 7 steps in [M&R, section 9-1.6] ().
Extract from [M&R, section 9-3.1, section 8-2.2] ().

0. * Assumptions *: Data $ x_1, \ dotsc, x_n $ is sample of $ \ Normal {\ mu, \ sigma} $ distributed probability variables, with $ \ sigma $ unknown.

1. * Parameter of interest *: Expectation value $ \ mu $.

2. * Null hypothesis *: For a fixed value $ \ mu_0 $:
$$ H_0: \ mu = \ mu_0 \,. $$

3. * Alternative hypothesis *:
$$ H_1: \ mu \ not = \ mu_0 \,. $$

4. * Test variable *:

Repetition: The sample mean (or the empirical mean) is
$$ \ Mean {x}: = \ frac {1} {n} \ sum_ {i = 1} ^ n x_i $$
and is the empirical standard deviation
$$ s: = \ sqrt {\ frac {1} {n-1} \ sum_ {i = 1} ^ n (x_i- \ Mean {x}) ^ 2} \,. $$
The test statistic is
$$ t: = \ frac {\ Mean {x} - \ mu_0} {s / \ sqrt {n}} \,. $$


5. * Decision *:
Choose significance level $ \ alpha \ in [0,1] $.
Reject $ H_0 $ as the $ p $ value $ \ le \ alpha $.

6. * Calculation *:
If $ T \ DistAs \ TDist {n-1} $, one calculates the $ p $ value
$$ p: = \ Proba (T \ le - \ Modulus {t} \ text {or} \ Modulus {t} \ le T)
= 2 \ Proba (T \ ge \ Modulus {t})
= 2 \ Proba (T \ le - \ Modulus {t}) \,. $$
See also [M&R, Appendix A, Table V] ().

7. * Conclusion *:
If $ p \ ge {} \ alpha $, then you must continue to accept $ H_0 $.
If $ p \ le {} \ alpha $, you can reject $ H_0 $.

8. * Confidence interval *:
The confidence interval at the confidence level is $ 1- \ alpha $ on a scale of $ t $ below $ H_0 $
$$ \ left [\ Quantile \ left (\ frac {\ alpha} {2} \ right), \ Quantile \ left (1- \ frac {\ alpha} {2} \ right) \ right] \ ,, $$
where $ \ Quantile $ is the tilt function of a $ \ TDist {n-1} $ distribution.
Calculating back after the $ \ mu $ scale gives the confidence interval
$$ \ left [
  \ Mean {x} - \ frac {s} {\ sqrt {n}} \ Quantile \ left (1- \ frac {\ alpha} {2} \ right)
 , \ Mean {x} - \ frac {s} {\ sqrt {n}} \ Quantile \ left (\ frac {\ alpha} {2} \ right)
  \ right] \,. $$
  
## Data
`` `{r}
data <- load.mr.mnemonic ('M&R 9-57 guinea pig birth weight in g')
`` `
`` `{r}
summary
sd (data)
length (data)
`` `
`` `{r}
hist (data)
`` `

## Review problem

See [M&R, problem 9-57] ().

`` `{r}
alpha <- 0.05
mu.zero <- 310
`` `

$ H_0: \ mu = 310 $.

## Manual calculation

`` `{r}
n <- length (data)
v <- n - 1
m <- mean (data)
s <- sd (data)
t <- (m-mu.zero) / s * sqrt (n)
t
p <- 2 * pt (-abs (t), df = v)
p
p <= alpha
`` `

Key with test value `r t` and $ p $ value` r p`.
Because `p <= alpha` is equal to` r p <= alpha`, we continue to accept $ H_0 $.

`` `{r}
interval <- m - s / sqrt (n) * c (qt (1 - alpha / 2, v), qt (alpha / 2, v))
interval
`` `

The $ `r 100 * (1 - alpha)` $% confidence interval for $ \ mu_0 $ is $ [`r interval`] $.

`` `{r}
dom <- seq (-4, 4, length.out = 100)
plot (dom, pt (dom, v), type = 'l', col = 'black', main = 'P-value calculation')
lines (dom, dt (dom, f), col = 'blue')

t.left <- abs (t)
p.left <- pt (t.left, v)
lines (c (-4.4), c (p.left, p.left), col = 'green')
dom2 <- seq (-4, t.left, length.out = 20)
polygon (c (-4, dom2, t.left), c (0, dt (dom2, v), 0), col = 'red', border = NA)

t.right <- abs (t)
p.right <- pt (t.right, v, lower.tail = F)
lines (c (-4.4), 1 - c (p.right, p.right), col = 'green')
dom3 <- seq (t.right, 4, length.out = 20)
polygon (c (t.right, dom3, 4), c (0, dt (dom3, v), 0), col = 'red', border = NA)
`` `

## Calculation with R
`` `{r}
key <- t.test (data, mu = mu.zero, alternative = "two.sided", conf.level = 1 - alpha)
test
`` `

`` `{r}
key $ statistic
`` `