---
title: "Chi-kwadraat homogeneity-test"
author: "Rob van der Nagel"
date: "12-3-19"
header-includes:
  - \input{lib/macros.tex}
output:
  html_notebook:
    df_print: paged
    toc: yes
    toc_depth: 3
    toc_float: true
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: 3
urlcolor: blue
---

```{r, include = F}
rm(list = ls())
source('lib/utils.r')
```
```{r, child = c('lib/macros.tex')}
```

# Motivation homogeneity test

Two samples from a discrete distribution. Are these samples independently distributed? Usually used to show that there is a connection (but not, of what kind).

# Theory homogeneity test

Application of Pearson test from [M&R, section 9-7] ().

I. * Assumptions *:
$ x_1, \ dotsc, x_n $ sample from $ P $ on $ \ Set {1, \ dotsc, k_1} \ times \ Set {1, \ dotsc, k_2} $.

II. * Greatness of interest *:
Distribution $ P $.
Edge divisions $ P_1 $ on $ \ Set {1, \ dotsc, k_1} $ and $ P_2 $ on $ \ Set {1, \ dotsc, k_2} $.

III. * Hypotheses *:
$$ H_0: P = P_1 \ otimes P_2 \. $$
$$ H_0: P \ not = P_1 \ otimes P_2 \. $$

IV. * Test statistic *:

The * empirical frequency * counts how many data are in each course:
$$
  f:
  \ qquad
  \ Set {1, \ dotsc, k_1} \ times \ Set {1, \ dotsc, k_2} \ to \ Nat
  \ qquad
  (i, j) \ mapsto {} \ Cardinality {\ Set {1 \ le l \ le n \ mid x_l = (i, j)}}
  \ ,.
$$
The * empirical fringe frequencies * are
$$
  f ^ {(1)}:
  \ qquad
  \ Set {1, \ dotsc, k_1} \ to \ Nat
  \ qquad
  i \ mapsto {} \ Cardinality {\ Set {1 \ le l \ le n \ mid x_l = (i ,.)}}
$$
and
$$
  f ^ {(2)}:
  \ qquad
  \ Set {1, \ dotsc, k_2} \ to \ Nat
  \ qquad
  j \ mapsto {} \ Cardinality {\ Set {1 \ le l \ le n \ mid x_l = (., j)}}
  \ ,.
$$
We compare the empirical frequency with the product of the peripheral frequencies.
The estimator for the frequency of product distribution of the edge distributions is
$$
  f ^ \ otimes:
  \ qquad
  \ Set {1, \ dotsc, k_1} \ times \ Set {1, \ dotsc, k_2} \ to \ Nat
  \ qquad
  (i, j) \ mapsto {} \ frac {f ^ {(1)} (i) f ^ {(2)} (j)} {n}
  \ ,.
$$
The test statistic is
$$
  k ^ 2: =
   \ sum_ {i = 1} ^ {k_1} \ sum_ {j = 1} ^ {k_2}
   \ frac {\ Bigl (f (i, j) -f ^ \ otimes (i, j) \ Bigr) ^ 2} {f ^ \ otimes (i, j)}
  \ ,.
$$
For sufficiently large $ n $, $ K ^ 2 $ has a $ \ ChiSquare {(k_1 - 1) (k_2 - 1)} $ distribution.
In practice, the central limit theorem requires $ f ^ {1} (i) f ^ 2 (j) $ not to be too small (rule of thumb: $ f ^ {1} (i) f ^ 2 (j) {} \ 5 $ for each course).

V. * Decision *:
Choose a significance level $ \ alpha \ in [0,1] $.
Reject $ H_0 $ as the $ p $ value $ \ le \ alpha $.

VI. * Confidence interval *:
If $ K ^ 2 \ DistAs \ ChiSquare {k-1} $, the $ p $ value is calculated by
$$ p: = \ Proba (k ^ 2 \ le K ^ 2) \,. $$
See also [M&R, Appendix A, Table IV] ().


`` `{r}
<- load.mr.mnemonic ('M&R 9-103 machine breakdown per shift')
fall out
`` `
`` `{r}
chisq.test (drop out)
`` `
We may not reject "$ H_0 $: shift is machine independent" at significance level $ 5 $%.

# Example: two dice

`` `{r}
d1 <- sample (1: 6, 200, replace = TRUE)
 # + 1 mod 6
d2 <- c (6, 1, 2, 3, 4, 5) [d1]
d3 <- sample (1: 6, 200, replace = TRUE)
# split and reverse magnitude within halfs
d4 <- c (3, 2, 1, 6, 5, 4) [d3]
d5 <- (d1 + d4) %% 6 + 1
d6 <- pmax (d2, d3)

dice <- data.frame (d1, d2, d3, d4, d5, d6)
`` `

`` `{r}
library ('corrplot')
corrplot :: corrplot (cor (dice))
`` `


`` `{r}
pairs (dice)
`` `

Cross table between d1 and d6 (2d frequency table):
`` `{r}
t.16 <- table (d1, d6)
`` `
`` `{r}
t.16
addmargins (t.16)
addmargins (prop.table (t.16))
`` `

A scatter plot does not properly show how many discrete values ​​there are, only where they are.
`` `{r}
plot (d1, d6)
`` `

A quick way to display a crosstab.
`` `{r}
# https://stackoverflow.com/questions/3712402/r-how-to-change-lattice-levelplot-color-theme#3712855
lattice :: level plot (t.16
                 , col.regions = gray (200: 100/200)
                 , main = 'D1 vs D6'
                 )
`` `

`` `{r}
# https://stackoverflow.com/questions/20673584/visualizing-crosstab-tables-with-a-plot-in-r#20674019
library ('ggplot2')
g <- ggplot (as.data.frame (t.16), aes (d1, d6)) + geom_point (aes (size = Freq), color = "green") + theme_bw () + xlab ("") + ylab ("")
g + scale_size_continuous (range = c (0, 20)) + geom_text (aes (label = Freq))
`` `

Two forms of the same test.
`` `{r}
chisq.test (d1, d6)
`` `
`` `{r}
chisq.test (t.16)
`` `
The degrees of freedom are $ 25 = (6-1) ^ 2 $.
We conclude that `d1` and` d6` are not independent at all.

`` `{r}
chisq.test (d1, d4)
`` `
We may not reject "$ H_0 $:` d1` and `d4` are independent".

If one wants to test on a certain connection, then a Pearson test with a given $ 2d $ distribution must be made.

`` `{r}
table (d1, d6)
`` `

`` `{r}
n <- sum (t.16)
P.0 <matrix (
  c (0, 0, 0, 0, 0, 6,
    1, 1, 1, 1, 1, 1,
    0, 2, 1, 1, 1, 1,
    0, 0, 3, 1, 1, 1,
    0, 0, 0, 4, 1, 1,
    0, 0, 0, 0, 5, 1
    ) / 36
  nrow = 6, ncol = 6, byrow = T)
P.0
sum (P.0)
`` `
`` `{r}
axis.vector (t.16)
F.0 <- P.0 * n
axis.vector (F.0)
chisq.test (as.vector (t.16), p = as.vector (F.0), rescale.p = T)
`` `
Here we have $ 35 $ instead of just $ 25 $ degrees of freedom, because we don't test against product distribution.
Problem with key: too many $ 0 $ boxes.
We correct by filtering.
`` `{r}
chisq.test (t.16 [F.0> = 5], p = F.0 [F.0> = 5], rescale.p = T)
`` `
With $ 20 $ degrees of freedom remaining, we continue to accept $ H_0: P = P_0 $.

# Other graphs for cross tables

`` `{r}
# https://stackoverflow.com/questions/33634593/define-breaks-for-hist2d-in-r
library ('gplots')
gplots :: hist2d (d1, d6, nbins = c (6, 6))
`` `

`` `{r}
library ('descr')
t <- descr :: CrossTable (d1, d6)
plot (t)
`` `

`` `{r}
mosaic plot (t.16, color = T)
`` `

`` `{r}
library ('vcd')
vcd :: mosaic (t.16)
`` `

`` `{r}
heatmap (t.16, Rowv = NA, Colv = NA, col = terrain.colors (diff (range (t.16))), scale = NULL)
`` `

`` `{r}
# heatmap.2 (t.16, scale = "none", dendrogram = "none", trace = "none", density.info = "histogram")
`` `

`` `{r}
library ('RColorBrewer')
num.colors <- 8
colors <- RColorBrewer :: brewer.pal (num.colors, "Pastel1")
t.16.cut <- cut (t.16, breaks = num.colors)
t.16.cut
levels (t.16.cut) [t.16.cut]
t.16.recoded <- matrix (unclass (t.16.cut), nrow = 6, ncol = 6, byrow = TRUE)
t.16.recoded
heatmap (t.16.recoded, Rowv = NA, Colv = NA, col = RColorBrewer :: brewer.pal (num.colors, "Pastel1"))
`` `
